---
title: "Annotate antibody structural contacts"
author: "Tyler Starr"
date: "10/11/2020"
output:
  github_document:
    toc: true
    html_preview: false
editor_options:
  chunk_output_type: inline
---
This notebook analyzes antibody-bound RBD crystal and cryo-EM structures to annotate structural contacts. It generates a csv listing structurally-defined contact residues for each ligand (i.e. Ab or ACE2).

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","bio3d","tidyverse")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#read in config file for determining which structure and chains to determine as contacts
contacts_config <- read_yaml(file=config$structural_contacts_config)

#make output directory
if(!file.exists(config$structural_contacts_dir)){
  dir.create(file.path(config$structural_contacts_dir))
}
```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

## Annotate sites on the RBD that contact antibody

We use the `binding.site` function from the `bio3d` package to identify residues that are structural complexes within each pdb. We iterate through the values in `structural_annotation_config.yaml` to perform this calculation for each structure described by that configuration file.

### Important note:
Chains cannot be called `N` in the config file. This will lead to an error (read as not a character). 
Change all `N` chains to `"N"`.

```{r annotate_contacts}
#empty data frame to append contacts to
structural_contacts <- data.frame(name=character(0), pdb=character(0), chain=character(0),position=numeric(0))
antibody_contacts <- data.frame(name=character(0), pdb=character(0), chain=character(0), position=numeric(0))

for(entry in contacts_config){
  pdb <- read.pdb(file=entry$pdbfile)
  # get the name of the pdb file to be include in output CSV file
  pdb_short <- strsplit(entry$pdbfile, split = "/") %>% 
    unlist() %>% tail(n=1) %>% strsplit("\\.") %>% unlist() %>% head(n=1)
  contacts <- binding.site(pdb,
                         a.inds=atom.select(pdb,chain=entry$chains_RBD),
                         b.inds=atom.select(pdb,chain=entry$chains_ligand),
                         cutoff=entry$distance_cutoff, hydrogens=F)
  if (is.vector(contacts)){
    structural_contacts <- rbind(structural_contacts, 
                                 data.frame(name=entry$name, 
                                            pdb=pdb_short, 
                                            chain = contacts$chain, 
                                            position = contacts$resno)
                                 )
  }

}
```

## Annotate sites on the antibody that contact the RBD
This is slightly more complicated because each antibody has multiple chains. We examine each chain individually, and output a separate CSV file that includes the chain designation.
In addition, there are some antibody chains that do not contact the RBD at all, so we must deal with this edge case.

## IMPORTANT DISCLAIMER!!!!
Importantly, some PDB files (such as is the case for the CR3022 antibody, PDB 6W41) annotate the CDR loops with letters that are stripped when read by the `bio3d` package so you should check each PDB file individually to make sure this is not the case, otherwise there will be problems with this script.
```{r annotate_antibody_contacts}
# make a separate file for the sites on the antibody that contact RBD
for(entry in contacts_config){
  pdb <- read.pdb(file=entry$pdbfile)
  # get the name of the pdb file to be include in output CSV file
  pdb_short <- strsplit(entry$pdbfile, split = "/") %>% 
    unlist() %>% tail(n=1) %>% strsplit("\\.") %>% unlist() %>% head(n=1)
  contacts <- binding.site(pdb,
                         a.inds=atom.select(pdb,chain=entry$chains_ligand),
                         b.inds=atom.select(pdb,chain=entry$chains_RBD),
                         cutoff=entry$distance_cutoff, hydrogens=F)
  if (is.vector(contacts)){
    antibody_contacts <- rbind(antibody_contacts, 
                               data.frame(name=entry$name, 
                                          pdb=pdb_short, 
                                          chain = contacts$chain, 
                                          position = contacts$resno))

  }

 }

```
## Save list of structural contacts

```{r output_structural_contacts}
write.csv(structural_contacts,file=config$structural_contacts,row.names=F,quote=F)
write.csv(antibody_contacts,file=config$antibody_contacts,row.names=F,quote=F)
```
